Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-01-23 doc/invt.tex
small changes to invt.tex

\documentclass[reprint, superscriptaddress, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[table,usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{setspace}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
often the task is to compute
a distribution, $p^*(z)$,
along a collective variable $z$ by sampling via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The logarithm, $\ln p^*(z)$,
defines a free energy, or a potential of mean force (PMF).
%
Difficulty arises, however, when $z$
represents a collective variable
that tends to move locally in a multi-modal distribution.
%
Thus,
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF, resulting sampling of a flat distribution
along $z$\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%The negative logarithm,
%$-\ln p^*(z)$,
%defines a free energy,
%or the potential of mean force (PMF).
%%
%Normally, the distribution $p^*(z)$ is
%limited in one or a few narrow $z$ regions.
%%
%Therefore, to capture the global shape
%of the PMF,
%one has to run multiple independent simulations
%under different umbrella or bias potentials
%along $z$\cite{mezei1987, berg1992, lee1993},
%and then piece together the entire curve.
%%
%This solution can be laborious % clumsy in practice
%and it often fails to converge quickly
%for a complex and glassy system,
%because the system can be trapped
%in a local free energy minimum.
%%for a long time.
%
%
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, lee1993,
%wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
Both techniques employ regular updating of the bias potential
to discourage future visits to the current configuration
by incrementally elevating the bias potential there.
%
The difference lies
in the updating window function,
which takes the form of a discrete
$\delta$-function (or a boxcar function one histogram bin wide)
in the WL case,
or that of a Gaussian
in the metadynamics.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{laio2002},
and for convergence
one  has to gradually decrease
the updating magnitude.

The schedule of reducing
the updating magnitude,
therefore, critically affects
the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal schedule
is a function of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
or
the inverse number of simulation steps.
%
The optimal schedule for the metadynamics case,
however, is less studied.

@@

The second type of methods,
which we will refer to as the adaptive FDS methods,
address the above shortcomings
by continuously and incrementally update the target distribution.
%For this reason, the second type of methods,
%referred to as the adaptive FDS methods,
%were invented.
%
Examples include
the WL algorithm and metadynamics.
%
%These methods incrementally update the bias potential
%and can approximately sample the target distribution.
%
However,
%they are less well understood theoretically, as
as mentioned, the on-the-fly updates break the microscopic reversibility
of the underlying sampling mechanism.
Further, failure to properly reduce the magnitude of
the histogram-based increments
would lead to roughness or errors in the final estimate.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods adopt a schedule to
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
%Thus, in late stages of a long simulation,
%the updating becomes negligible and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
%if the bias potential were converged to the exact value,
%this adaptive FDS tends to the ideal equilibrium counterpart,
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
%
For example, one can show that
if the magnitude decreases according to
the optimal inverse-time prescription\cite{
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014},
the WL algorithm is equivalent to the ideal equilibrium counterpart
in improving the precision of the bias potential,
delivering the maximal efficiency
(cf. Appendix \ref{sec:equilerr}).
%the bias potential in the WL algorithm
%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%with the precision of the histogram as simulation lengthens.

@@

We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where the elements, $w_{ij}$'s, are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used in metadynamics
is a multiple-bin scheme.

@@

In the original WL papers\cite{
wang2001, wang2001pre},
the updating magnitude $\alpha$
(or $\ln f$ there)
is controlled in stages.
%
In each stage, $\alpha$
is kept constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha$ reduced by a factor of $1/2$\cite{
wang2001, wang2001pre}.
%
While this scheme works well for early stages,
%
it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
Here we will re-derive this formula
and suggest a slight modification
whereby the origin of $t$ is shifted
according to the initial error.
%
Interestingly, we may reach the same result
from optimally planning the simulation lengths in the stages
and the reduction factor, as shown in Appendix \ref{sec:wltoinvt}.

@@

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution $\pi_i(t)$,
%
the real deviation is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
In this way,
the variables, $x_i(t)$'s,
represent the error of the bias potential,
and are expected to vanish on convergence.
%
We, therefore, define
the error of the bias potential as
%
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n \rho_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n \rho_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}

@@

For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_V}
for the deterministic part
%
\begin{align}
  \langle h_i(t) \rangle
  &\approx
  \pi_i(t)
  =
  \frac{                p_i \, e^{-x_i(t)} }
       { \sum_{j = 1}^n p_j \, e^{-x_j(t)} }
  \notag
  \\
  &\approx
  p_i \,
  \left(
    1 - x_i + \sum_{r=1}^n p_r \, x_r
  \right)
  ,
  \label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $x_i(t)$ in the linear expansion
and used $\sum_{j=1}^n p_j = 1$
on the second line.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - x_i  }
     { \sum_{ r = 1 }^n p_r (1 - x_r) }
=
\frac{                       1 - x_i  }
     { 1 - \sum_{ r = 1 }^n p_r \, x_r }
.
$$
}%

@@

The above characterization allows us to determine
the optimal initial updating magnitude.
%
Let us now consider simulations in which
the system is initially equilibrated at the same
constant updating magnitude $a_0$.
%
For these simulations,
we can adjust $q(T)$
to minimize the total error.
%
From Eqs. \eqref{eq:error_res1}
and \eqref{eq:error_asym2},
we find that
the $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, (a_0 / 2) }
  .
\label{eq:opt_qT}
\end{equation}
%
\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial E_R
  }
  {
    \partial Q_c
  }
  &=
  -a_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial E_A
  }
  {
    \partial Q_c
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (E_R + E_A) / \partial Q_c = 0$.
}%
By Eqs. \eqref{eq:mQ_invTa} and \eqref{eq:intmQ_tT},
we can rewrite the condition as
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e., the optimal initial updating magnitude
is always half of the equilibration value.
%
This factor $1/2$
happens to coincide with the
recommended reduction factor of updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.

@@

Parallel to Eq. \eqref{eq:phi_pbc},
we define the out-of-boundary values
of $\phi_i$ as
%
\begin{equation}
  \phi_i
  =
  \begin{dcases}
    \phi_{ 1 - i }           & \mathrm{for \;} i \le 0, \\
    \phi_{ 2 \, n + 1 - i }  & \mathrm{for \;} i > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that the matrix multiplication is still equivalent to
a convolution as Eq. \eqref{eq:wmul_to_convol}.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}%
We note that Eq. \eqref{eq:phi_refl}
defines a periodic variable of period $2 \, n$
with a reflective symmetry around $i = 1/2$.
%
Accordingly, the eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
must share the same periodicity,
and be even functions about the axis $i = 1/2$:
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, 2 \, \pi}
            {                    2 \, n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
The eigenvalues can be borrowed from
Eq. \eqref{eq:wband_eigenvalue_pbc},
with the periodic $n$ doubled:
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \, 2 \, \pi}{ 2 \, n }
  .
\label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{Derivation.
  First consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies Eq. \eqref{eq:phi_refl}.
  %
  So, by Eq. \eqref{eq:wmul_to_convol},
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue_refl}.
}%
%
We may unify Eqs. \eqref{eq:wband_eigenvalue_refl}
and \eqref{eq:wband_eigenvalue_pbc} as
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2
  \sum_{ l = 1 }^b
    \mu_l \,
    \cos \frac{ (k - 1) \, l \, g \, \pi }
              {            n             }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where the degeneracy $g$ is $2$
for the periodic case,
or $1$ otherwise.
%
In both cases,
the eigenvalues, $\lambda_k$'s,
are the cosine transform of
the updating kernel.
%
For the updating scheme to be stable,
the cosine transform of the kernel
must be nonnegative.
%
If this is not so,
we can modify the kernel
using the technique described
in Appendix \ref{sec:stabilize_wband}.

@@

We thank Dr. J. Ma, Dr. Y. Mei, J. Wang,
O. Nassar, Dr. C. Lai, Dr. S. Ou, D. Stuhlsatz,
Dr. C. Myers, and Dr. O. Samoylova
for helpful discussions.  We gratefully acknowledge the Robert A. Welch Foundation (H-0037), the National Science Foundation (CHE-1152876) and the National Institutes of Health (GM-037657) for partial support of this work.  This research used computing resources of the National Science Foundation XSEDE grid.

>>>>

\documentclass[preprint, superscriptaddress, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[table,usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{setspace}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
often the task is to compute
a distribution, $p^*(z)$,
along a collective variable $z$ by sampling via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The logarithm, $\ln p^*(z)$,
defines a free energy, or a potential of mean force (PMF).
%
Difficulty arises, however, when $z$
represents a collective variable
that tends to move locally in a multi-modal distribution.
%
Thus,
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF, resulting sampling of a flat distribution
along $z$\cite{mezei1987, berg1992, lee1993,
wang2001, *wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%The negative logarithm,
%$-\ln p^*(z)$,
%defines a free energy,
%or the potential of mean force (PMF).
%%
%Normally, the distribution $p^*(z)$ is
%limited in one or a few narrow $z$ regions.
%%
%Therefore, to capture the global shape
%of the PMF,
%one has to run multiple independent simulations
%under different umbrella or bias potentials
%along $z$\cite{mezei1987, berg1992, lee1993},
%and then piece together the entire curve.
%%
%This solution can be laborious % clumsy in practice
%and it often fails to converge quickly
%for a complex and glassy system,
%because the system can be trapped
%in a local free energy minimum.
%%for a long time.
%
%
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, lee1993,
%wang2001, *wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, *wang2001pre}
and metadynamics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
Both techniques employ regular updating of the bias potential
to discourage future visits to the current configuration
by incrementally elevating the bias potential there.
%
The difference lies
in the updating window function,
which takes the form of a discrete
$\delta$-function (or a boxcar function one histogram bin wide)
in the WL case,
or that of a Gaussian
in the metadynamics.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{laio2002},
and for convergence
one  has to gradually decrease
the updating magnitude.

The schedule of reducing
the updating magnitude,
therefore, critically affects
the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, *belardinelli2007jcp, *belardinelli2016, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, *caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal schedule
is a function of the
inverse time\cite{liang2007,
belardinelli2007, *belardinelli2007jcp, *belardinelli2016, belardinelli2008,
morozov2007, zhou2008}
or
the inverse number of simulation steps.
%
The optimal schedule for the metadynamics case,
however, is less studied.

@@

The second type of methods,
which we will refer to as the adaptive FDS methods,
address the above shortcomings
by continuously and incrementally update the target distribution.
%For this reason, the second type of methods,
%referred to as the adaptive FDS methods,
%were invented.
%
Examples include
the WL algorithm and metadynamics.
%
%These methods incrementally update the bias potential
%and can approximately sample the target distribution.
%
However,
%they are less well understood theoretically, as
as mentioned, the on-the-fly updates break the microscopic reversibility
of the underlying sampling mechanism.
Further, failure to properly reduce the magnitude of
the histogram-based increments
would lead to roughness or errors in the final estimate.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods adopt a schedule to
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, *belardinelli2007jcp, *belardinelli2016, belardinelli2008,
barducci2008}.
%
%Thus, in late stages of a long simulation,
%the updating becomes negligible and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
%if the bias potential were converged to the exact value,
%this adaptive FDS tends to the ideal equilibrium counterpart,
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
%
For example, one can show that
if the magnitude decreases according to
the optimal inverse-time prescription\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2016, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, *caparica2014},
the WL algorithm is equivalent to the ideal equilibrium counterpart
in improving the precision of the bias potential,
delivering the maximal efficiency
(cf. Appendix \ref{sec:equilerr}).
%the bias potential in the WL algorithm
%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%with the precision of the histogram as simulation lengthens.

@@

We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where the elements, $w_{ij}$'s, are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used in metadynamics
is a multiple-bin scheme.

@@

In the original WL papers\cite{
wang2001, *wang2001pre},
the updating magnitude $\alpha$
(or $\ln f$ there)
is controlled in stages.
%
In each stage, $\alpha$
is kept constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha$ reduced by a factor of $1/2$\cite{
wang2001, *wang2001pre}.
%
While this scheme works well for early stages,
%
it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2016, belardinelli2008}.

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2016, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, *caparica2014}.
%
Here we will re-derive this formula
and suggest a slight modification
whereby the origin of $t$ is shifted
according to the initial error.
%
Interestingly, we may reach the same result
from optimally planning the simulation lengths in the stages
and the reduction factor, as shown in Appendix \ref{sec:wltoinvt}.

@@

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution $\pi_i(t)$,
%
the real deviation is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
In this way,
the variables, $x_i(t)$'s,
represent the error of the bias potential,
and are expected to vanish on convergence.
%
We, therefore, define
the error of the bias potential as
%
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n \rho_i \, \left\langle x_i^2(t) \right\rangle
  .
\label{eq:error_sum}
\end{equation}

@@

For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_V}
for the deterministic part
%
\begin{align}
  \langle h_i(t) \rangle
  &\approx
  \pi_i(t)
  =
  \frac{                p_i \, e^{-x_i(t)} }
       { \sum_{j = 1}^n p_j \, e^{-x_j(t)} }
  \notag
  \\
  &\approx
  p_i \,
  \left[
    1 - x_i(t) + \sum_{r=1}^n p_r \, x_r(t)
  \right]
  ,
  \label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $x_i(t)$ in the linear expansion
and used $\sum_{j=1}^n p_j = 1$
on the second line.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - x_i  }
     { \sum_{ r = 1 }^n p_r (1 - x_r) }
=
\frac{                       1 - x_i  }
     { 1 - \sum_{ r = 1 }^n p_r \, x_r }
.
$$
}%

@@

The above characterization allows us to determine
the optimal initial updating magnitude.
%
Let us now consider simulations in which
the system is initially equilibrated at the same
constant updating magnitude $a_0$.
%
For these simulations,
we can adjust $q(T)$
to minimize the total error.
%
From Eqs. \eqref{eq:error_res1}
and \eqref{eq:error_asym2},
we find that
the $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, (a_0 / 2) }
  .
\label{eq:opt_qT}
\end{equation}
%
\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial E_R
  }
  {
    \partial Q_c
  }
  &=
  -a_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial E_A
  }
  {
    \partial Q_c
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (E_R + E_A) / \partial Q_c = 0$.
}%
By Eqs. \eqref{eq:mQ_invTa} and \eqref{eq:intmQ_tT},
we can rewrite the condition as
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e., the optimal initial updating magnitude
is always half of the equilibration value.
%
This factor $1/2$
happens to coincide with the
recommended reduction factor of updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, *wang2001pre}.

@@

Parallel to Eq. \eqref{eq:phi_pbc},
we define the out-of-boundary values
of $\phi_i$ as
%
\begin{equation}
  \phi_i
  =
  \begin{dcases}
    \phi_{ 1 - i }           & \mathrm{for \;} i \le 0, \\
    \phi_{ 2 \, n + 1 - i }  & \mathrm{for \;} i > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that the matrix multiplication is still equivalent to
a convolution as Eq. \eqref{eq:wmul_to_convol}.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}%
We note that Eq. \eqref{eq:phi_refl}
defines a periodic variable of period $2 \, n$
with a reflective symmetry around $i = 1/2$.
%
Accordingly, the eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
must share the same periodicity,
and be even functions about the axis $i = 1/2$:
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, 2 \, \pi}
            {                    2 \, n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
The eigenvalues can be borrowed from
Eq. \eqref{eq:wband_eigenvalue_pbc},
with the period $n$ doubled:
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \, 2 \, \pi}{ 2 \, n }
  .
\label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{Derivation.
  First consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies Eq. \eqref{eq:phi_refl}.
  %
  So, by Eq. \eqref{eq:wmul_to_convol},
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue_refl}.
}%
%
We may unify Eqs. \eqref{eq:wband_eigenvalue_refl}
and \eqref{eq:wband_eigenvalue_pbc} as
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2
  \sum_{ l = 1 }^b
    \mu_l \,
    \cos \frac{ (k - 1) \, l \, g \, \pi }
              {            n             }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where the degeneracy $g$ is $2$
for the periodic case,
or $1$ otherwise.
%
In both cases,
the eigenvalues, $\lambda_k$'s,
are the cosine transform of
the updating kernel.
%
For the updating scheme to be stable,
the cosine transform of the kernel
must be nonnegative.
%
If this is not so,
we can modify the kernel
using the technique described
in Appendix \ref{sec:stabilize_wband}.

@@

We thank Dr. Y. Mei, J. Wang,
O. Nassar, Dr. C. Lai, Dr. S. Ou, D. Stuhlsatz,
Dr. C. Myers, and Dr. O. Samoylova
for helpful discussions.  We gratefully acknowledge the Robert A. Welch Foundation (H-0037), the National Science Foundation (CHE-1152876) and the National Institutes of Health (GM-037657) for partial support of this work.  This research used computing resources of the National Science Foundation XSEDE grid.
